---
title: "Programming in R"
author: "John Hinic"
date: '2022-07-02'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "images/")
```

## R Programming Reflections

Well, the summer session only has about 1 month left at this point, which feels kind of insane. Regardless, at this point we are done formally learning about "how to program" in R. I think that, for the most part, most of the material would be considered review for me - some areas I absolutely needed the review, while others served more as helpful practice. There were also some topics that we went a good bit deeper than what I knew coming into the class.

One example would be pretty much the whole concept of "functional programming." I've learned about writing functions and their benefits before, and even use little helper functions a fair amount (my job also uses SAS, and I use macros quite a bit for that). But I feel like I've started to scratch the surface of *real* functional programming, which is really exciting and cool on the one hand, but also pretty overwhelming on the other.

When considering the things I am most and least prepared for in industry, I think the process of actually putting a model into production is one of the big ones I will need to learn more about. From what I understand, one part of this (from a data scientist's perspective) is essentially just wrapping everything up in functions and putting them in scripts, so it's been really cool learning more about functions in general.

### Example

The general rule of thumb is that if you ever write the same code more than twice, you should put it into a function instead. One time I find myself needing to copy paste similar code a lot is when I'm performing some EDA and want to make a bunch of plots with different groupings or different aesthetics. So, to demonstrate the usefulness of writing functions to perform these repetitive tasks, I'm going to write a function that allows me to create a range of plots using `ggplot2` and the `mpg` dataset.

Before starting, we'll load in the `tidyverse` library and show the data we'll be using.

```{r}
library(tidyverse)
head(mpg)
```

For the sake of this demonstration, let's assume the only numeric variable we are interested in is `hwy`, and we want to make both box plots and histograms of that variable. We'll write a function that takes in the type of plot we want ("box" or "hist"), if/how we want to group it (only one level of grouping), and how we want to display the groups ("color", "facet", or "both").

```{r function}
plotting <- function(type, grp = NULL, grpDisp = "facet") {
  plotfun <- switch(
    type,
    "box" = geom_boxplot,
    "hist" = geom_histogram,
    stop("Invalid plot type supplied")
  )
  if(!is.null(grp)){
    if(grpDisp %in% c("color", "both")) aesOpt <- paste0(", fill = ", grp)
    else aesOpt <- ""
    
    if(grpDisp %in% c("facet", "both")) facet <- paste0('+ facet_wrap(~', grp, ')')
    else facet <- ""
  } else {
      aesOpt <- ""
      facet <- ""
  }
  
  funcStr <- paste0('mpg %>% ', paste0('ggplot(aes(', paste0('hwy', aesOpt), ')) + plotfun()', facet))
  funcCall <- str2expression(funcStr)
  eval(funcCall)
}
```

Now, we can call this single function with only a few arguments to create plots instead of writing several lines for each one. For example:

```{r plotting}
plotting("hist")
plotting("hist", grp = "drv")
plotting("hist", grp = "drv", grpDisp = "both")
plotting("box", grp = "manufacturer", grpDisp = "color")
plotting("box", grp = "trans", grpDisp = "both")
```

This can speed up our workflow quite a bit as long as we have a solid understanding of how to write the functions in a reasonable time-frame.